Integrating Big Data and Deep Learning for Improved Wheat (Triticum aestivum) Grain Yield Predictions

### 2023-05-23 submission_informatika.tex
added tables and figures

@@

\[
y = \mu + G\tau + Eu + e,
\]

@@

Similar to the phenotypic data, the genotypic data for six experimental series was inherited from \cite{zhao2021unlocking} and additional genotypic data was added to supplement the phenotypic data. In total this data holds the genetic information for ~ 14k genotypes spread across nine chips of varying marker densities and genotype overlap (Table 1). A reference chip (chip 9) was selected since it had maximum marker overlap with other chips and its maker's oligoes were retrieved from the original publication \cite{wang2014characterization}. The oligos were then used to get the physical position of respective SNP variants on Chinese spring v1 release of wheat genome (ref?). The reference and alternative SNP calls at the reference chip were used to judge if a marker was inverted in a given chip or not. And the pipeline (Fig. 1) was followed to integrate the arrays.

@@

The environment relationship matrix (\textit{ERM}) was then calculated as:
\[
\textit{ERM}_l = EC \cdot EC^T
\]
\[
\textit{ERM}_nl = \exp\left(\frac{{\text{dist}(EC)}}{\theta}\right)
\]
where EC is the scaled Environment covariate matrix, $EC^T$ is the transpose of the EC matrix, $\text{dist}(EC)$ is the Euclidean Distance matrix of EC, and $\theta$ is a scaling factor.

@@

The robustness of genotypic data integration was assessed by examining the pairwise $F_{st}$ statistic between major population groups and utilizing the integrated data for genomic predictions \cite{ref?} of those groups. The Exp\textunderscore5 population, which included hybrids and parents from a wide cross of elite lines with genebank material, served as an outgroup (Figure \ref{fig3}). The observed genetic bifurcation aligns with biological expectations, as the remaining population groups primarily consisted of elite lines and hybrids resulting from crosses between elite lines.

@@

The performance of E-GBLUP and dynamic CNN models was benchmarked using three cross-validation scenarios. In the first scenario, traditional cross-validation was employed, where genotypes were randomly divided into a training and a test population in an 80:20 ratio. CNN model outperformed E-GBLUP for predicting line grain yield (Figure \ref{fig4}). For the second scenario, stratified cross-validation was implemented to account for genetic distinctness observed in one experimental series (Figure \ref{fig4}). The results showed slightly higher prediction accuracy in some cases compared to traditional cross-validation, but the overall trend of higher performance in case of lines was preserved here. The third scenario aimed to investigate the effect of utilizing BigData on prediction accuracy. Six scenarios were designed with gradually increasing training population sizes. The results demonstrated that as the training population size increased, the prediction accuracy improved, highlighting the utility of BigData for genomic prediction (Figure \ref{fig5}). Overall, Convolutional Neural Networks (CNN) exhibited superior performance compared to the traditional GBLUP based method, especially when dealing with extensive data or when flexibility in the genetic structure of training and test populations was required.

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\graphicspath{ {./pictures/} }

@@

\vspace{6pt}

\hspace{0.5cm}\begin{minipage}{\dimexpr\linewidth-0.5cm}
\begin{equation}
y = \mu + G\tau + Eu + e,
\end{equation}
\end{minipage}

\vspace{6pt}

@@

Similar to the phenotypic data, the genotypic data for six experimental series was inherited from \cite{zhao2021unlocking} and additional genotypic data was added to supplement the phenotypic data. In total this data holds the genetic information for ~ 14k genotypes spread across nine chips of varying marker densities (Table \ref {chip_overview}) and genotype overlap (Table \ref {chip_overlap}). A reference chip (chip 9) was selected since it had maximum marker overlap with other chips and its maker's oligoes were retrieved from the original publication \cite{wang2014characterization}. The oligos were then used to get the physical position of respective SNP variants on Chinese spring v1 release of wheat genome (ref?). The reference and alternative SNP calls at the reference chip were used to judge if a marker was inverted in a given chip or not. And the pipeline  (Fig. \ref{fig_pipeline_marker}) was followed to integrate the arrays.

\begin{table}[htbp]
\centering
%\setlength{\tabcolsep}{1pt}
\caption{Overview of genotypic data derived from nine sources i.e. chip\textunderscore1 to chip\textunderscore9. The second and the third column shows the absolute number of genotypes and markers in any given chip.}
\label{chip_overview}
\begin{tabular}{ l c c c l }
 \hline
 chip{\textunderscore}name & Genotypes & Marker{\textunderscore}unique \\
 \hline
 chip\textunderscore1  & 901    & 21062 \\
 chip\textunderscore2  & 725    & 20821 \\
 chip\textunderscore3  & 155    & 35865 \\
 chip\textunderscore4  & 145    & 81489 \\
 chip\textunderscore5  & 1579   & 75761 \\
 chip\textunderscore6  & 1707   & 14411 \\
 chip\textunderscore7  & 968    & 11736 \\
 chip\textunderscore8  & 1396   & 13006 \\
 chip\textunderscore9  & 371    & 81587 \\
 \hline
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\setlength{\tabcolsep}{5pt}
\caption{Overview of genotypic data overlap derived from nine sources i.e. chip\textunderscore1 to chip\textunderscore9.  The upper and lower triangular half of the columns 2 to 10 shows the marker and genotype overlap between pairs of sources.}
\label{chip_overlap}
\begin{tabular}{ l c c c c c c c c c c l }
 \hline
 chip{\textunderscore}name  & chip\textunderscore1 & chip\textunderscore2 & chip\textunderscore3 & chip\textunderscore4 & chip\textunderscore5 & chip\textunderscore6 & chip\textunderscore7 & chip\textunderscore8 & chip\textunderscore9  \\
 \hline
 chip\textunderscore1   & NA      & 20821   & 14245   & 17835   & 16264   & 13584   & 11264   & 12552   & 17313\\
 chip\textunderscore2   & 0       & NA      & 14245   & 17594   & 16264   & 13345   & 11264   & 12552   & 17313\\
 chip\textunderscore3   & 0       & 0       & NA      & 34394   & 32729   & 11496   & 9769    & 10796   & 35865\\
 chip\textunderscore4   & 0       & 0       & 0       & NA      & 75676   & 14388   & 11380   & 12275   & 77977\\
 chip\textunderscore5   & 3       & 0       & 2       & 0       & NA      & 13000   & 11157   & 11603   & 75731\\
 chip\textunderscore6   & 4       & 0       & 1       & 0       & 167     & NA      & 10814   & 11897   & 13735\\
 chip\textunderscore7   & 3       & 0       & 1       & 0       & 48      & 71      & NA      & 10221   & 11730\\
 chip\textunderscore8   & 2       & 0       & 6       & 0       & 15      & 13      & 18      & NA      & 12896\\
 chip\textunderscore9   & 3       & 0       & 2       & 0       & 12      & 7       & 8       & 22      & NA   \\
 \hline
\end{tabular}
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{pipeline_marker}
    \caption{Outlines of the steps in merging the chips. Four chips colored grey to yellow are used for representation. For a subset of markers (indicated by red box), each marker was evaluated based on overlapping genotypes to correct for markers inversions and genotyping errors.}
    \label{fig_pipeline_marker}
\end{figure}

@@

The environment relationship matrix (\textit{ERM}) was then calculated as:

\vspace{6pt}

\begin{equation}
\hspace{0.5cm}\textit{ERM}_l = EC \cdot EC^T
\end{equation}

\vspace{6pt}

\begin{equation}
\hspace{0.5cm}\textit{ERM}_{nl} = \exp\left(\frac{{\text{dist}(EC)}}{\theta}\right)
\end{equation}

\vspace{6pt}

where EC is the scaled Environment covariate matrix, $EC^T$ is the transpose of the EC matrix, $\text{dist}(EC)$ is the Euclidean Distance matrix of EC, and $\theta$ is a scaling factor.

@@

The robustness of genotypic data integration was assessed by examining the pairwise $F_{st}$ statistic between major population groups and utilizing the integrated data for genomic predictions of those groups. The Exp\textunderscore5 population, which included hybrids and parents from a wide cross of elite lines with genebank material, served as an outgroup (Fig. \ref{fig_fst_plot}). The observed genetic bifurcation aligns with biological expectations, as the remaining population groups primarily consisted of elite lines and hybrids resulting from crosses between elite lines.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{fst_plot}
    \caption{$F_{st}$ statistic calculated between population groups indicate genetic divergence between the groups. The dendrogram was constructed by applying multidimensional scaling on pairwise $F_{st}$ matrix to show clustering between recently developed (outlined by the red box) and PGR derived population groups.}
    \label{fig_fst_plot}
\end{figure}

@@

The performance of E-GBLUP and dynamic CNN models was benchmarked using three cross-validation scenarios. In the first scenario, traditional cross-validation was employed, where genotypes were randomly divided into a training and a test population in an 80:20 ratio. CNN model outperformed E-GBLUP for predicting line grain yield (Fig. \ref{fig_pred_corr_res_NN_2}). For the second scenario, stratified cross-validation was implemented to account for genetic distinctness observed in one experimental series (Fig. \ref{fig_pred_corr_res_NN_2}). The results showed slightly higher prediction accuracy in some cases compared to traditional cross-validation, but the overall trend of higher performance in case of lines was preserved here. The third scenario aimed to investigate the effect of utilizing BigData on prediction accuracy. Six scenarios were designed with gradually increasing training population sizes. The results demonstrated that as the training population size increased, the prediction accuracy improved, highlighting the utility of BigData for genomic prediction (Figure \ref{fig5}). Overall, Convolutional Neural Networks (CNN) exhibited superior performance compared to the traditional GBLUP based method, especially when dealing with extensive data or when flexibility in the genetic structure of training and test populations was required.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{pred_corr_res_NN_2}
    \caption{Genomic prediction accuracy (corr on y-axis between 0 to 1 ) of GBLUP and Convolutional Neural Networks (NN) for mean grain yield across sites under 5-fold cross validation (cv) and stratified cross validation (st). Results are shown separately for hybrids and lines.}
    \label{fig_pred_corr_res_NN_2}
\end{figure}
